A portable, annotation-based, visual stepper for Common Lisp

### 2020-04-06 paper/paper.tex
Final paper version

By the same token, if one does implement such a tool, it should be
done in a manner that is of actual, practical use.  This was the
thinking behind the SLY/Emacs stepper.  We think it especially
fortunate that annotation-based approach and the manner of source-code
correlation in SLY/Emacs's stickers feature doesn't make use of direct
references to source files or file positions.  As was shown in
\ref{stickers}, we merely keep an enumerated list of stickers
identifiers synchronized between Common Lisp and SLY/Emacs, a
mechanism that is simple but effective.  It adequately resists some
modifications to the source of the instrumented form or its
whereabouts, such as moving it around in the source file, or even
adding white-space and comments inside it.  This detail is crucial in
making stickers and stepping actually usable for day-to-day
programming, since the user isn't dragged away from his editor or
forced to a special confinement while stepping.  We note that these
advantages of annotation-based steppers were already hinted at back in
1989 by the proponents of evaluation-based steppers\cite[l.41,
l.42]{evaluation-based}.

By contrast, a hypothetical evaluation-based visual stepper (such an
as enhanced version of \cite{bourguignon}) would find it difficult to
maintain this advantage since the source-code correlation is achieved
at evaluation-time and is harder to mutate effectively afterwards.
Perhaps this fact can explain why the non-portable visual steppers of
the Allegro CL and LispWorks implementations don't work directly with
the source-editing facilities present in these
IDEs\footnote{LispWorks's stepper manual specifically mentions this
  fact in \cite{lispworks-stepper}.}.  A possible solution would be to
represent source-code correlation in terms of sub-expression paths in
the form tree instead of character positions, but it would still be
hard to resist deletions and insertions at top level.  A more
heavy-handed solution would lock the source file read-only for the
duration of the stepping session.  However, users are normally adverse
to such confinements.

We also think it fortunate that stepping is implemented as an
increment to the existing stickers functionality.  As in
\cite{annotation-based}, we consider it an advantage of the
annotation-based stepping systems that users are allowed to instrument
only the definitions they are interested in.  Indeed, it is often the
case that stepper systems become tedious to operate because they step
on \emph{too much}.  Our system has an additional selectivity
advantage: since the annotations are stickers, users can manually
adjust the automatically placed stickers, removing the ones they are
not interested in, or adding others.

It may also be noted that the usual stepping paradigm where the
program is stopped at each point is only one of the possibilities
afforded by stickers.  As we explained in sections \ref{stickers} and
\ref{results}, two further ones can constitute innovative means of
debugging: the \emph{post-mortem} replay of sticker recordings and the
visual decoration of source code with colors to indicate the most
recent state of each sticker's execution.

@@

Its current limitations notwithstanding, the SLY/Emacs stepper
described in this essay is an effective example of a portable, visual
stepping facility for the Common Lisp ecosystem.  To the best of our
knowledge, it is indeed the \emph{only} system that combines these
characteristics.  As such, there is little to compare it against,
except maybe for the aforementioned \texttt{edebug.el}\cite{edebug}
program for the Emacs Lisp language, an annotation-based stepper that
handles macros via user-supplied syntax declarations.  We believe that
the underpinnings of the SLY/Emacs stepper -- \emph{stickers} and
mnesic macroexpansion -- are solid.  We envision enhancements to its
user interface, certainly by incorporating ideas from the interfaces
of non-portable visual Common Lisp steppers, or steppers for other
programming languages.

>>>>

By the same token, if one does implement such a tool, it should be
done in a manner that is of actual, practical use.  This was the
reasoning behind the SLY/Emacs stepper.  We think it especially
fortunate that the annotation-based approach and the manner of
source-code correlation in SLY/Emacs's stickers don't make use of
direct references to source files or file positions.  As was shown in
\ref{stickers}, we merely keep an enumerated list of stickers
identifiers synchronized between Common Lisp and SLY/Emacs, a
mechanism that is simple but effective.  It adequately resists some
modifications to the source of the instrumented form or its
whereabouts, such as moving it around in the source file, or even
adding white-space and comments inside it.  This detail is crucial in
making stickers and stepping usable for day-to-day programming, since
the user isn't dragged away from his editor or forced to a special
confinement while stepping.  We note that these advantages of
annotation-based steppers were already hinted at back in 1989 by the
proponents of evaluation-based steppers\cite[l.41,
l.42]{evaluation-based}.

By contrast, a hypothetical evaluation-based visual stepper (such an
as enhanced version of \cite{bourguignon}) would find it difficult to
maintain this advantage since the source-code correlation is achieved
at evaluation-time and is harder to mutate effectively afterwards.
Perhaps this fact can explain why the non-portable visual steppers of
the Allegro CL and LispWorks implementations don't work directly with
the source-editing facilities present in these IDEs.  A possible
solution would be to represent source-code correlation in terms of
sub-expression paths in the form tree instead of character positions,
but it would still be hard to resist deletions and insertions at top
level.  A more heavy-handed solution would lock the source file
read-only for the duration of the stepping session.  However, users
are normally adverse to such confinements.

We also think it fortunate that stepping is implemented as an
increment to the existing stickers functionality.  As in
\cite{annotation-based}, we consider it an advantage of the
annotation-based systems that users are allowed to instrument only the
definitions they are interested in.  Indeed, it is often the case that
steppers become tedious to operate because they step on \emph{too
  much}.  Yet, in our system, users can manually adjust the
automatically placed stickers, removing the ones they are not
interested in, or adding others.

It may also be noted that the usual stepping paradigm where the
program is stopped at each point is only one of the possibilities
afforded by stickers.  As we explained in sections \ref{stickers} and
\ref{results}, two further ones constitute innovative means of
debugging: the \emph{post-mortem} replay of sticker recordings and the
visual decoration of source code with colors indicating the state of
the most recent execution of each sticker.

@@

The SLY/Emacs stepper described in this essay is an effective example
of a portable, visual stepping facility for the Common Lisp ecosystem.
To the best of our knowledge, it is indeed the \emph{only} system
combining these characteristics.  As such, there is little to compare
it against.  The closest match could well lie outside of Common Lisp,
in the aforementioned \texttt{edebug.el}\cite{edebug} stepper, an
annotation-based approach that is equally well integrated with the
source code editor.  That stepper has a better interface, but requires
custom declarations for stepping into macro expansions, something the
SLY/Emacs stepper handles automatically.  Its current limitations
notwithstanding, we believe that the underpinnings of the SLY/Emacs
stepper -- \emph{stickers}, a simple source-tracking reader, and
mnesic macroexpansion -- are solid.  We envision enhancements to its
interface, perhaps by incorporating ideas of non-portable visual
steppers, or steppers for other programming languages.
